Storm surges in the Strait of Georgia simulated with a regional model

### 2015-06-17 AO_guides/manuscriptAO.tex
corrected mean error calculations
final look through of mansucript. Improved clairty, corrected typos

{\color{red} Although Tofino is not the closest tide gauge to the western boundary, we found that the surge conditions on the western coast of Vancouver Island and near the mouth of the Strait of Juan de Fuca are fairly uniform as summarized in Table \ref{tab:pacificsurge}. Generally, surge amplitudes at Tofino are slightly larger and earlier than Neah Bay and Bamfield. The remote forcing is applied uniformly across the open boundaries. Given the similarities between the surge amplitude and phase at Neah Bay and Bamfield, this seems to be a reasonable assumption. Further, a geostrophic slope calculation with velocity 20 cm s$^{-1}$ would imply a difference of less than 5 cm between the northern and southern sides of the Strait of Juan de Fuca. In the future, this uniform specification of sea level at the western boundary will be relaxed by taking values from the solution of a large-scale forecasting model.}

\begin{table}[h]
\centering 
\tbl{{\color{red}Average surge amplitudes and delay in peak timing from Tofino at several tide gauges on the Pacific coast. Averages obtained over the five surge events discussed on this article. A negative delay indicates that the peak surge occurred later than at Tofino.}} 
{\begin{tabular}{|l |c | c|}
\hline 
Location & Mean surge amplitude (m) & Mean delay from Tofino (hrs) \\
\hline
Port Hardy & 0.491 & -2.67\\
Winter Harbour & 0.572 & 0.333 \\
Tofino  & 0.707 & -- \\
Bamfield & 0.658 & -0.5 \\
Neah Bay & 0.658 & -1.83 \\
Toke Point & 1.17 & -2.83 \\
\hline
\end{tabular}}
\label{tab:pacificsurge}
\end{table}

@@

{\color{red} The ocean surface is forced with momentum fluxes, solar and non-solar heat fluxes, and precipitation from a 33-km global atmospheric reforecasting model suitable for use in ocean modelling \citep{smith2014new}.} Forecasts from the period of 2002-2012 are available. One additional simulation employs the western Canada component of the High Resolution Deterministic Prediction System (HRDPS), a nested 2.5 km resolution atmospheric model provided by Environment Canada \citep{ECModel}. In addition to momentum and heat fluxes, forcing due to the surface atmospheric pressure is also included in the ocean momentum equations \citep{madec2012nemo}. In the open ocean, the surface atmospheric pressure primarily induces an inverse barometer (IB) sea surface height,
\begin{equation}
 \eta_{IB} = -\frac{1}{g\rho_{0}}\left(P_{s}-P_0\right), \label{eq:inverse}
\end{equation}
where $g$ is the acceleration due to gravity, $\rho_{0}$ is a reference density set to 1035 kg m$^{-3}$, $P_{s}$ is the sea level atmospheric pressure from the atmospheric model, and $P_0$ is a reference sea level pressure (set to 101,000 Pa). Since the atmospheric model employs a terrain following vertical coordinate the model surface pressure has been corrected to sea level using \citep{holton1992introduction}
\[ P_s = P_1\left(\gamma\frac{z_1}{T_1} +1 \right)^\frac{g}{\gamma R},\]
where $P_1$ and $T_1$ are the model pressure and temperature at height $z_1$, $R$ is the ideal gas constant, and $\gamma$ is a temperature lapse rate set to 0.0098 K m$^{-1}$. In coastal and shelf seas, the atmospheric pressure forcing can also induce significant non-IB sea surface height, as demonstrated in previous studies of the Hudson Bay and Labrador shelf system \citep{wright1987influence,young1995synoptic}.

@@

Point Atkinson & 11.6&	3.401&	3.483&	0.013&	0.995&	6.7&	0.302&	0.307	&0.192	&0.932\\
Victoria       &10.8&	2.175	&2.162	&0.04	&0.989	&5.1	&0.29	&0.28&	0.167&	0.94\\
Patricia Bay   &11.6&	2.573&	2.619&	0.028&	0.991&	6.0&	0.281&	0.279&	0.179&	0.932\\
Campbell River & 11.6&	3.139&	3.177&	0.021&	0.994&	7.8&	0.246&	0.293&	0.188&	0.893\\

@@

\tbl{{\color{red}Summary of maximum water level, maximum surge amplitude and model delay for all simulations. The model time delay is the number of hours between the maximum in the observations and the maximum in the model. A negative number means the model's maximum was later than the observations. Values calculated over the same periods discussed in Table \ref{tab:statistics}. } }

@@

Point Atkinson  & 5.196 & 5.277 & 0  & 0.752 & 0.634 & -3.125\\
Victoria        & 3.147 & 3.125 & -0.625 & 0.684 & 0.563 & -2.125\\
Patricia Bay    & 4.005 & 3.901 & -0.125 & 0.689 & 0.584 & -2.875\\
Campbell River  & 4.756 & 4.835 & -0.250 & 0.673 & 0.590 & -3.875\\ 

@@

{\color{red} This case was also used to assess the importance of the northern boundary residual forcing and briefly examine the effects of stratification. When the northern boundary forcing is excluded the surge amplitude at Campbell River drops by about 10 cm. The other stations were not strongly affected. This was also observed in another hindcast (Dec 2006). The effect of stratification was examined by considering a case with an initially horizontally uniform stratification and comparing to the fully stratified case. The largest differences were observed at Victoria, with an average 5 cm increase in sea surface height after an initial adjustment period. Other studies have seen an improvement in modelled surge amplitudes when using a baroclinic model as opposed to a three-dimensional barotropic model \citep{ma2015oceanic}. We plan to investigate stratified effects in more detail in a future study.}

@@

Hindcasts with both atmospheric forcing products agree well with the observations (Figure \ref{fig:dec2012_surge}). Willmott skill scores for both modelled residuals are above {\color{red} 0.84} at these four locations. {\color{red} There is considerable noise in the observed Campbell River residual, which could explain its low skill score (Table \ref{tab:statistics}).} {\color{red} The modelled residuals in each simulation deviate only slightly from each other, most noticeable at Point Atkinson during the strong wind event on Dec 15, 2012 where simulation with the low resolution atmospheric forcing has a residual that is 5 cm higher.  The maximum residuals on Dec 17, 2012 are almost identical at each station (Table \ref{tab:peak}).} Although the wind speeds on Dec 15, 2012 are quite high, this event did not result in any flooding or residuals greater than 50 cm. Significant flooding occurred later, on Dec 17, 2012, where the observed surge amplitude at Point Atkinson was {\color{red}66.3 cm}. Even though the low resolution model significantly under predicts the wind speed at some locations, it performs comparably to forcing with high resolution products. Much of the fine scale detail offered by the high resolution atmospheric model is averaged out since, to first order, the storm surge is a spatial integral of the wind. This result further demonstrates that the local wind is not the most important factor for setting up storm surges over this domain; however, local wind patterns contribute to wave conditions which can impact flooding and are not accounted for in this model.

@@

Comparisons between high resolution and low resolution atmospheric forcing further suggest that local winds are not the leading order factor in setting up large surge amplitudes. The low resolution model significantly underestimated wind speeds in certain areas of the domain, however, it performed just as accurately in reproducing surge amplitudes as the high resolution product.  Although the high resolution model delivers fine scale detail and more accurate local wind patterns, it did not provide any improvement to the surge hindcast. Since both of these products provide the same level of accuracy for storm surge hindcasts, in the future we will use the high resolution atmospheric model to produce real-time storm surge predictions, even though most of our testing was performed with the low resolution model. {\color{red} Additionally, the sea surface height anomaly boundary condition of the present model will be taken from the solution of large-scale forecasting model.}

>>>>

{\color{red} Although Tofino is not the closest tide gauge to the western boundary, we found that the surge conditions on the western coast of Vancouver Island and near the mouth of the Strait of Juan de Fuca are fairly uniform as summarized in Table \ref{tab:pacificsurge}. Generally, surge amplitudes at Tofino are slightly larger and earlier than Neah Bay and Bamfield. The remote forcing is applied uniformly across the open boundaries. Given the similarities between the surge amplitude and phase at Neah Bay and Bamfield, this seems to be a reasonable assumption. In support of this assumption, a geostrophic slope calculation with an estuarine velocity of 20 cm s$^{-1}$ would imply a difference of less than 5 cm between the northern and southern sides of the Strait of Juan de Fuca. In the future, this uniform specification of sea level at the western boundary will be relaxed by taking values from the solution of a large-scale forecasting model.}

\begin{table}[h]
\centering 
\tbl{{\color{red}Average peak surge amplitudes and delay in peak timing from Tofino at several tide gauges on the Pacific coast. Averages obtained over the five surge events hindcasted in this article. A negative delay indicates that the peak surge occurred later than at Tofino.}} 
{\begin{tabular}{|l |c | c|}
\hline 
Location & Mean surge amplitude (m) & Mean delay from Tofino (hrs) \\
\hline
Port Hardy & 0.491 & -2.67\\
Winter Harbour & 0.572 & 0.333 \\
Tofino  & 0.707 & -- \\
Bamfield & 0.658 & -0.5 \\
Neah Bay & 0.658 & -1.83 \\
Toke Point & 1.17 & -2.83 \\
\hline
\end{tabular}}
\label{tab:pacificsurge}
\end{table}

@@

{\color{red} The ocean surface is forced with momentum fluxes, solar and non-solar heat fluxes, and precipitation from a 33-km global atmospheric reforecasting model {\color{red}called CGRF} suitable for use in ocean modelling \citep{smith2014new}.} Forecasts from the period of 2002-2012 are available. One additional simulation employs the western Canada component of the High Resolution Deterministic Prediction System (HRDPS), a nested 2.5 km resolution atmospheric model provided by Environment Canada \citep{ECModel}. In addition to momentum and heat fluxes, forcing due to the surface atmospheric pressure is also included in the ocean momentum equations \citep{madec2012nemo}. In the open ocean, the surface atmospheric pressure primarily induces an inverse barometer (IB) sea surface height,
\begin{equation}
 \eta_{IB} = -\frac{1}{g\rho_{0}}\left(P_{s}-P_0\right), \label{eq:inverse}
\end{equation}
where $g$ is the acceleration due to gravity, $\rho_{0}$ is a reference density set to 1035 kg m$^{-3}$, $P_{s}$ is the sea level atmospheric pressure from the atmospheric model, and $P_0$ is a reference sea level pressure (set to 101,000 Pa). Since the atmospheric model employs a terrain following vertical coordinate the model surface pressure has been corrected to sea level using \citep{holton1992introduction}
\[ P_s = P_1\left(\gamma\frac{z_1}{T_1} +1 \right)^\frac{g}{\gamma R},\]
where $P_1$ and $T_1$ are the model pressure and temperature at height $z_1$, $R$ is the ideal gas constant, and $\gamma$ is a temperature lapse rate set to 0.0098 K m$^{-1}$. In coastal and shelf seas, the atmospheric pressure forcing can also induce significant non-IB sea surface height, as demonstrated in previous studies of the Hudson Bay and Labrador shelf system \citep{wright1987influence,young1995synoptic}.

@@

Point Atkinson & 12.0 & 3.424 & 3.514 & 0.017 & 0.993 & 7.2 & 0.318 & 0.327 & 0.176 & 0.933\\
Victoria       & 9.6 & 2.193 & 2.193 & 0.042 & 0.988 & 5.3 & 0.295 & 0.295 & 0.162 & 0.945\\
Patricia Bay   & 11.0 & 2.587 & 2.648 & 0.030 & 0.989 & 6.6 & 0.285 & 0.293 & 0.175 & 0.932\\
Campbell River & 11.4 & 3.184 & 3.219 & 0.026 & 0.993 & 7.4 & 0.280 & 0.311 & 0.172 & 0.920\\

@@

\tbl{{\color{red}Summary of maximum water level, maximum surge amplitude and model delay for all simulations. The model time delay is the number of hours between the maximum in the observations and the maximum in the model. A negative number means the model's maximum was later than the observations. Values are calculated over the same periods discussed in Table \ref{tab:statistics}. } }

@@

Point Atkinson & 5.106 & 5.194 & 0 & 0.806 & 0.681 & -2.6 \\
Victoria       & 3.148 & 3.130 & -1.0 & 0.696 & 0.589 & -1.0\\
Patricia Bay   & 3.939 & 3.856 & -0.2 & 0.724 & 0.614 & -2.2\\
Campbell River & 4.726 & 4.796 & -0.4 & 0.730 & 0.640 & -2.6 \\

@@

{\color{red} This case was also used to assess the importance of the northern boundary residual forcing and to briefly examine the effects of stratification. When the northern boundary forcing is excluded the surge amplitude at Campbell River drops by about 10 cm. The other stations were not strongly affected. This was also observed in another hindcast (Dec 2006). The effect of stratification was examined by considering a case with an initially horizontally uniform stratification and comparing to the fully stratified case. The largest differences were observed at Victoria, with an average 5 cm increase in sea surface height after an initial adjustment period. Other studies have seen an improvement in modelled surge amplitudes when using a baroclinic model as opposed to a three-dimensional barotropic model \citep{ma2015oceanic}. We plan to investigate stratified effects in more detail in a future study.}

@@

Hindcasts with both atmospheric forcing products agree well with the observations (Figure \ref{fig:dec2012_surge}). Willmott skill scores for both modelled residuals are above {\color{red} 0.84} at these four locations. {\color{red} There is considerable noise in the observed Campbell River residual, which could explain its low skill score (Table \ref{tab:statistics}).} {\color{red} The modelled residuals in each simulation deviate only slightly from each other, most noticeable at Point Atkinson during the strong wind event on Dec 15, 2012 where the simulation with the low resolution atmospheric forcing has a residual that is 5 cm higher.  The maximum residuals on Dec 17, 2012 are almost identical at each station (Table \ref{tab:peak}).} Although the wind speeds on Dec 15, 2012 are quite high, this event did not result in any flooding or residuals greater than 50 cm. Significant flooding occurred later, on Dec 17, 2012, where the observed surge amplitude at Point Atkinson was {\color{red}66.3 cm}. Even though the low resolution model significantly under predicts the wind speed at some locations, it performs comparably to forcing with high resolution products. Much of the fine scale detail offered by the high resolution atmospheric model is averaged out since, to first order, the storm surge is a spatial integral of the wind. This result further demonstrates that the local wind is not the most important factor for setting up storm surges over this domain; however, local wind patterns contribute to wave conditions which can impact flooding and are not accounted for in this model.

@@

Comparisons between high resolution and low resolution atmospheric forcing further suggest that local winds are not the leading order factor in setting up large surge amplitudes. The low resolution model significantly underestimated wind speeds in certain areas of the domain; however, it performed just as accurately in reproducing surge amplitudes as the high resolution product.  Although the high resolution model delivers fine scale detail and more accurate local wind patterns, it did not provide any improvement to the surge hindcast. Since both of these products provide the same level of accuracy for storm surge hindcasts, in the future we will use the high resolution atmospheric model to produce real-time storm surge predictions, even though most of our testing was performed with the low resolution model. {\color{red} Additionally, the sea surface height anomaly boundary condition of the present model will be taken from the solution of large-scale forecasting model.}
